---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PostCard

<!-- badges: start -->
[![R-CMD-check](https://github.com/NNpackages/PostCard/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/NNpackages/PostCard/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/NNpackages/PostCard/graph/badge.svg)](https://app.codecov.io/gh/NNpackages/PostCard)
<!-- badges: end -->

PostCard provides tools for accurately estimating marginal effects using GLMs, including increasing precision using prognostic covariate adjustment (see introductory examples below, more details in `rctglm()` and )

**Marginal effects** are causal effects of the form $r(\Psi_1, \Psi_0)$, where $\Psi_a=\mathbb{E}[Y(a)]$ are population mean outcomes under exposure $a=0, 1$, respectively. These are sometimes referred to as *counterfactual means*.

The package uses **plug-in estimation for robust estimation of any marginal effect estimand** as well as **influence functions for robust estimation of the variance of the estimand** (Rosenblum, M. and M. J. van der Laan, 2010: Simple, efficient estimators of treatment effects in randomized trials using generalized linear models to leverage baseline variables. The International Journal of Biostatistics, 6, no. 1).

## Installation

You can install the development version of PostCard from [GitHub](https://github.com/) with:

``` r
pak::pak("NNpackages/PostCard")
```
Setup-chunk to load the package, set a seed and turn off verbosity for the rendering of the README.

```{r setup, message = FALSE}
library(PostCard)
withr::local_seed(1395878)
withr::local_options(list(PostCard.verbose = 0))
```

# Plug-in estimation of marginal effects and variance estimation using influence functions

## Simulating data for exploratory analyses

First, we simulate some data to be able to enable showcasing of the functionalities. For this we use the `glm_data()` function from the package, where the user can specify an expression alongside variables and a family of the response to then simulate a response from a GLM with linear predictor given by the expression provided.

```{r dat-sim}
n <- 1000
b0 <- 1
b1 <- 3
b2 <- 2

# Simulate data with a non-linear effect
dat_treat <- glm_data(
  Y ~ b0+b1*sin(W)^2+b2*A,
  W = runif(n, min = -2, max = 2),
  A = rbinom(n, 1, prob = 1/2),
  family = gaussian() # Default value
)
```



### Fitting `rctglm()` without prognostic covariate adjustment

The `rctglm()` function estimates any specified estimand using plug-in estimation for randomised clinical trials and estimates the variance using the influence function of the marginal effect estimand. 

The interface of `rctglm()` is similar to that of the `stats::glm()` function but with an added mandatory specification of

-   The randomisation variable in data, usually being the (name of) the treatment variable
-   The randomisation ratio - the probability of being allocated to group 1 (rather than 0)
    -   As a default, a ratio of 1's in data is used
-   An estimand function
    -   As a default, the function takes the average treatment effect (ATE) as the estimand

Thus, we can estimate the ATE by simply writing the below:

> Note that as a default, `verbose = 2`, meaning that information about the algorithm is printed to the console. However, here we suppress this behavior. See more in `vignette("model-fit")`.

```{r noprog-run}
ate <- rctglm(formula = Y ~ A * W,
              exposure_indicator = A,
              exposure_prob = 1/2,
              data = dat_treat,
              family = "gaussian") # Default value
```

This creates an `rctglm` object which prints as
```{r noprog-show}
ate
```

The structure of such an `rctglm` object is broken down in the **`Value`** section of the documentation in `rctglm()`.
    
Methods available are `estimand` (or the shorthand `est`) which prints a `data.frame` with and estimate of the estimand and its standard error. A method for `coef` is also available to extract coefficients from the underlying `glm` fit.
```{r}
est(ate)
```

See more info in the documentation page `rctglm_methods()`.

### Using prognostic covariate adjustment

The `rctglm_with_prognosticscore()` function uses the `fit_best_learner()` function to fit a prognostic model to historical data and then uses the prognostic model to predict
\begin{align}
\mathbb{E}[Y|X,A=0]
\end{align}

for all observations in the current data set. These *prognostic scores* are then used as a covariate in the GLM when running `rctglm()`.

Allowing the use of complex non-linear models to create such a prognostic score allows utilising information from potentially many variables, "catching" non-linear relationships and then using all this information in the GLM model using a single covariate adjustment.

We simulate some historical data to showcase the use of this function as well:
```{r hist-data}
dat_notreat <- glm_data(
  Y ~ b0+b1*sin(W)^2,
  W = runif(n, min = -2, max = 2),
  family = gaussian # Default value
)
```

The call to `rctglm_with_prognosticscore()` is the same as to `rctglm()` but with an added specification of

-   (Historical) data to fit the prognostic model using `fit_best_learner()`
-   A formula used when fitting the prognostic model
    -   Default uses all covariates in the data.
-   (Optionally) number folds in cross validation and a list of learners for fitting the best learner

Thus, a simple call which estimates the average treatment effect, adjusting for a prognostic score, is seen below:
```{r prog-run}
ate_prog <- rctglm_with_prognosticscore(
  formula = Y ~ A * W,
  exposure_indicator = A,
  exposure_prob = 1/2,
  data = dat_treat,
  family = gaussian(link = "identity"), # Default value
  data_hist = dat_notreat)
```

Quick results of the fit can be seen by printing the object:
```{r prog-show}
ate_prog
```

It's evident that in this case where there is a non-linear relationship between the covariate we observe and the response, adjusting for the prognostic score reduces the standard error of our estimand approximation by quite a bit.

#### Investigating the prognostic model

Information on the prognostic model is available in the list element `prognostic_info`, which the method `prog()` can be used to extract. A breakdown of what this list includes, see the **`Value`** section of the `rctglm_with_prognosticscore()` documentation.
