---
title: "Controlling behavior of model fitting functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Controlling behavior of model fitting functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message = FALSE}
library(PostCard)
library(magrittr)
withr::local_seed(1395878)
```

For introductory examples on how to use the package, see the [README file](../README.md).

In this vignette, we will explain how to alter the default behavior of the model fitting functions `rctglm()` and `rctglm_with_prognosticscore()`.

## Controlling verbosity

See package level options documentation in `options()`, giving information on how to change package behavior through options and environmental variables. Only option is `verbose`, which controls the amount of information printed to the console.

As a default, `verbose = 2`, meaning various information printed throughout the algorithm. Change to `verbose = 1` for a little less information or `verbose = 0` for no information

> Throughout the examples in this vignette, it will be visible what kind of information is printed to the user.

## Plug-in estimation of marginal effects and variance estimation using influence functions

As in the [README](https://github.com/NNpackages/PostCard#README) examples, we simulate data using the `glm_data()` function from the package.

```{r dat-sim}
n <- 1000
b0 <- 1
b1 <- 3
b2 <- 2

# Simulate data with a non-linear effect
dat_pois <- glm_data(
  Y ~ b0+b1*sin(W)^2+b2*A,
  W = runif(n, min = -2, max = 2),
  A = rbinom(n, 1, 1/2),
  family = poisson(link = "log") # Default value
)
```

### Changing the estimand

The default `estimand_fun` in `rctglm()` and `rctglm_with_prognosticscore()` is the average treatment effect (ATE). However, it's possible to specify any estimand.

#### Built-in estimands - average treatment effect and rate ratio

Built in is the ATE and rate ratio, which can be specified with character strings. As is apparent from the documentation of `rctglm()` and `rctglm_with_prognosticscore()`, the default of `estimand_fun` is `"ate"`, and similarly the user can specify `estimand_fun = "rate_ratio"` to use the estimand function `psi1 / psi0` as seen below:

```{r rate-ratio-run-show}
rate_ratio <- rctglm(
  formula = Y ~ A + W,
  exposure_indicator = A,
  exposure_prob = 1/2,
  data = dat_pois,
  family = "poisson",
  estimand_fun = "rate_ratio")
rate_ratio
```

#### Specifying any estimand

It's possible to specify any function with 2 named arguments, `psi0` and `psi1`, as your estimand function `estimand_fun`. Below is an example showing that any estimand function can be specified.

> Note that information is printed to the console about the automatic differentiation being performed on the `estimand_fun`, which are used for the calculation of the influence function of the marginal effect.

```{r nonsense-run-show}
nonsense_estimand_fun <- function(psi1, psi0) {
  psi1 / sqrt(psi0) * 2 - 1
}

nonsense_estimand <- rctglm(
  formula = Y ~ A * W,
  exposure_indicator = A,
  exposure_prob = 1/2,
  data = dat_pois,
  family = poisson(),
  estimand_fun = nonsense_estimand_fun,
  verbose = 0)
nonsense_estimand
```

### Variance estimation using cross validation

The variance is estimated as the variance of the influence function of the marginal effect. During the calculation of this function, counterfactual prediction are made for all observations, using a GLM to predict their outcome in case they were in group 0 and 1, respectively. 

`cv_variance` is an argument in `rctglm` and `rctglm_with_prognosticscore` that enables obtaining these counterfactual predictions as out-of-sample (OOS) prediction by using cross validation. Using cross validation for the variance estimation is the default behavior in the package.

### Prognostic covariate adjustment

The `rctglm_with_prognosticscore()` uses the function `fit_best_learner()` to fit a prognostic model to the historical data, `data_hist`. Thereafter, the model is used to predict prognostic scores for all observations in `data` before using these scores as a covariate when performing plug-in etimation in a GLM using `rctglm`.

The default behavior of `fit_best_learner()` and subsequently `rctglm_with_prognosticscore()` is to fit a discrete super learner (discrete to avoid overfitting) by finding the model with the lowest RMSE among a linear regression, Multivariate adaptive regression spline (MARS) and boosted trees.

The algorithm uses a default of 5 folds for cross validation and tries to construct the formula used for the prognostic model as modelling the response with a design matrix containing an intercept and main effect from all variables in the historical data, `data_hist`.

#### Specifying learners

It's easy for the user to specify other learners to train in order to fit the discrete super learner. The package utilises the framework of [tidymodels](https://www.tidymodels.org/). All the user needs to do is specify a list of models.

##### Code used for default learners

To illustrate, below is the code of the unexported function `default_learners` which creates a list of the default learners:
```{r default-learners-show}
PostCard:::default_learners
```

##### Creating own list of learners

> We first simulate some data to run the example

```{r dat-hist-sim}
# Generate some data
n <- 1000
b0 <- 1
b1 <- 1.5
b2 <- 2
W_sim <- runif(n, min = -2, max = 2)

dat_treat <- glm_data(
  Y ~ b0+b1*abs(sin(W))+b2*A,
  W = W_sim,
  A = rbinom (n, 1, 1/2)
)

dat_notreat <- glm_data(
  Y ~ b0+b1*abs(sin(W)),
  W = W_sim
)
```

A listing of models is available at [the tidymodels website](https://www.tidymodels.org/find/parsnip/), and the user can specify a list of any of those models along with specified values of parameters to the `learners` argument.

```{r fit-own-learners}
learners <- list(
  rf = list(
    model = parsnip::rand_forest(
      mode = "regression",
      trees = 500,
      min_n = 5
    ) %>% 
      parsnip::set_engine("ranger")
  ),
  svm.linear = list(
    model = parsnip::svm_linear(
      mode = "regression",
      cost = 1,
      margin = 0.1) %>% 
    parsnip::set_engine("LiblineaR")
  )
)

model_own_learners <- rctglm_with_prognosticscore(
  formula = Y ~ A * W,
  exposure_indicator = A,
  exposure_prob = 1/2,
  data = dat_treat,
  family = gaussian(),
  data_hist = dat_notreat,
  learners = learners)
```

#### Investigating the prognostic model

It's possible to view information regarding the fit of the prognostic model in the `rctglm` class object that `rctglm_with_prognosticscore()` returns by looking at the list element `prognostic_info`.

Inside this list element are elements

- `model_fit`: The result of `fit_best_learner()`
- `learners`: The list of learners used
- `cv_folds`: The number of folds used for cross validation

```{r prognostic-info-show}
model_own_learners$prognostic_info
```


